---
title: "Quantitative Metazoan Metabarcoding"
author: "John Wares and Paula Pappalardo"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: pdf_document
bibliography: QuBar.bib
---

<!-- Paula: this manuscript is where I want to start clean from my prelim work to develop short paper together. -->
# Introduction

We start with what may seem like a trivial question: assume that you have been told that a series of fair coin flips resulted in 60% 'heads', 40% 'tails'. This is the only information given, but you already have made a judgment about how many coin flips occurred, and perhaps have generated a probability distribution in your head where the highest likelihood is for 5 or 10, rather than 50 or 100, events. This is taking advantage of what we know about the probability mass function of a binomial distribution, where the observed number of 'successes' in a series is related to the probability of success (presumably 50%) and the number of trials.  

Here, we argue that the same principle can be used for improving the efficiency of exploring the presence, distribution, and abundance of genetic biodiversity. Documenting the distribution and abundance of biodiversity - in many habitats, at multiple scales - is perhaps more important now than ever as scientists evaluate how populations are responding to environmental change. Though technological advances have rapidly improved some elements of this [@Nagendra01; @Bourlat13], there are still glaring deficiencies in our ability to efficiently catalog diversity, even in small domains or limited taxonomic surveys.

The most apparent advances have been in surveys of microbial and viral diversity. Next-generation sequencing has permitted the now-commonplace exploration of fungal, bacterial, and viral diversity by generating $10^5$ - $10^6$ sequence reads per sample and using barcoding approaches (match of sequence to known taxonomic samples for that genomic region) to identify the taxa present and their relative abundance. While there is no doubt that this has transformed our understanding of functional ecosystem processes at microbial ecology at this scale [@Nguyen14; @Turn09; @Desnues08], there are definite limitations. For example, some taxa (e.g. Archaea) may not be as readily amplified using the same ribosomal 16S "bacteria" primers, and variation in amplification efficiency certainly exists within the Eubacteria [@Acinas05]. Additionally, it is known that some bacterial genomes harbor more than one copy of this canonical locus [@Kembel12], thus muddling the relationship between read frequency and taxon frequency in a community.

The same problems exist - and are exacerbated - when studying multicellular diversity. Most notably, on top of the problems of potential contamination, detecting rare taxa and/or handling singleton evidence for rare taxa, and the potentially large variance in individual sizes of organisms, the relative read abundance in a NGS data set will often wildly vary (by multiple orders of magnitude) from the abundance of actual tissue in the data set [@Nguyen14; @Piñol14; @Bohmann14]. This is caused primarily by shifts in amplification efficiency given mismatches in the primer region, and is often dealt with by analyzing data for simple incidence as well as relative read abundance, to identify patterns robust to either removal of information or inaccurate information [@Nguyen14].

If, however, our goal is to understand the actual relative abundance of individuals of different species in a sample - with these species harboring variation at 'barcode' loci, and often being highly divergent from one another - the question is whether there is complementary information that can be extracted from these data that does not rely on the abundance of reads that are assigned to a taxon, but relies on our understanding of diversity within populations and how that can be measured. 

The summary statistics for DNA sequence diversity are well established and generally recognize the population mutation rate $\theta$ at a given locus; as a population increases in size, or as the mutation rate at that locus increases, more polymorphisms and more diversity will be found. There are limitations to this approach based on Kimura's neutral theory, as various forms of genomic selection will limit the direct relationship between population size and population diversity [e.g., @Bazin06; @Wares10; @Corbett15]. Nevertheless, these summary statistics - including Watterson's $\theta$, a sample-normalized estimator of $\theta$ using the number of segregating sites *S* in a sample - may provide information necessary to generate *some* information about abundance patterns from NGS data. This information also certainly has its limits: nucleotide diversity ($\pi$) requires information on polymorphic site frequencies that will be biased by differential amplification across individuals, as well as relatively uninformative - or diminishing returns - as the number of sampled individuals increases [@Wakeley08]. Haplotype diversity (*H*) is likely sufficient to set a minimum boundary on the number of individuals sampled, and *H* along with *S* have some information about the probability associated with larger numbers of individuals.

Here we present the mathematical considerations necessary to develop these quantitative tools, and then evaluate the situations in which there is sufficient power to make meaningful statements about relative abundance from polymorphism data alone. 

# Methods

```{r basic params, echo=FALSE,warning=FALSE,results='hide',message=FALSE,fig.show='asis'}

# you may have biomass and/or estimate number indivs, e.g. could see ~200 zooplankton in a sample
# put 20, 200, 2000 indivs on NGS to barcode will not change proportion of reads, just changes depths of coverage
# but this info constrains the posterior in units of individuals rather than proportions
# especially if there are multiple taxa, in which the minimum number of each can be subtracted to set upper boundaries

maxindivs=200
```

The approach here is identifying information that can comfortably be used as prior information to establish the posterior probability of observing polymorphism data from an *unknown* number of input individuals for a taxon. Some information is clearly limiting: for example, if it is known that only about `r maxindivs` individual specimens were originally used for isolation of DNA, then the maximum number of total individuals recovered from this approach should be about `r maxindivs`. This is perhaps not exciting numerical advance in biology, but limits our prior belief nonetheless.

```{r hapdiv, echo=FALSE,warning=FALSE,results='hide',message=FALSE,fig.show='asis'}

####################################
numhaps=10
Hapdiv=0.75 #minvalue 0.001 if no information; this is PRIOR information
####################################

x=1
cdf=0
indprob=0
array<-NULL
while (cdf<0.99) {
  cdfprev<-cdf
#  cdf<-pgamma(x,1,Hapdiv) 
#  if use 1 as shape parameter keeping shape parameter constant doesn't account for increased variance (?) as numhaps go up, 
#  e.g. error may be higher as you observe more...once it is working run it by somebody mathier.
  cdf<-pgamma(x,numhaps,Hapdiv) #might be that numhaps is actually the shape parameter!!!! or: something else...non-gamma.
  indprob<-cdf-cdfprev
  
  
  happrob<-numhaps+(x-1)
  array<-c(array,happrob)
  array<-c(array,cdf)
  array<-c(array,indprob)
#  print(happrob)
#  print(cdf)
  x=x+1
}

#par(mfrow=c(3,1))

probs<-t(matrix(array,nrow=3))
probs
#plot(probs[,1],probs[,3],col='red')

```

There are also clear minimum bounds that can be established for the abundance of a taxon. Considering DNA sequence haplotypes as our most basic information, we ask how many *distinct* haplotypes are recovered in the data that match a particular taxon? For a haploid mitochondrial marker like the oft-used cytochrome oxidase I (COI), this number is the minimum number of individuals present (if the number happens to be 0, it is also likely to be the maximum number of individuals in the sample). The number of haplotypes recovered from a sample of a species can be summarized with "haplotype diversity", *H*, defined by Nei and Tajima (1981) as $$ H = \frac{N}{N-1}(1-\sum\limits_{i=1}x_{i}^2) $$ and representing the probability that sampling a new individual will result in sample of a new haplotype. N is the number of haplotypes, and *x*~i~ is the sample frequency of the i~th~ haplotype.

An example of how *H* could be used is shown below for a sample in which `r numhaps` haplotypes are observed, *from a population or taxon with prior information about H*, here with *H* = `r Hapdiv`. In addition to assuming that prior information about the population is useful, here we assume a minimum of `r numhaps` individuals, and that what we do not know can be modeled by a Gamma distribution with the shape defined by the number of haplotypes and scale defined by our prior knowledge of *H*.

```{r fig1, echo=FALSE,warning=FALSE,results='hide',message=FALSE,fig.show='asis'}
plot(probs[,1],probs[,3],col='red')

```

So, observing `r numhaps` haplotypes for this taxon, we might feel comfortable believing there are between 16 and 31 actual individuals that were sampled. The problem lies in the willful abuse of the Gamma distribution without a better understanding of how haplotype diversity *H* and the sample size *N* are related through the frequency of haplotypes - remember, at this point we do not trust the proportion/frequency representation of an allele in our sample.

Another way to approach this is through 'true diversity' indices [reviewed in @Sherwin10], as this family of statistics indicates the number of equally abundant types that have an average frequency equal to the observed average of types. This would mean that instead of taking H as a given from a previously-studied taxon or population, the extant data would be used to calculate $$ ^qD = (\sum\limits_{i=1}^{R}p_{i}^q)^{1/(1-q)} $$ where *R* is richness (number of types). As the Simpson (1949) index is used to indicate the probability that two individuals taken at random from the data are of the same (haplo)type, a comparable statistic to *H* is reached by taking the inverse Simpson index, or $^2D$, and then transforming into the Gini-Simpson index as $1-1/^2D$, again the probability that two individuals have distinct haplotypes.

```{r NotoPrior, echo=FALSE,warning=FALSE,results='hide',message=FALSE,fig.show='asis'}
# rough estimates from Zakas Q is 10, hapdiv is 0.7
# grabbing a file from Geneious....
library(PopGenome)
file<-readData("FastaSeqs") # actual sample size here is 20. That is the number I'd like to come out...AND, TO AN EXTENT, IT IS AT LEAST IN THE INTERVALS FOR BOTH...
#actual<-20
file@n.sites
basic<-diversity.stats(file)
Hapdiv<-basic@hap.diversity.within #haplotype diversity

Nsite<-basic@n.biallelic.sites #number of seg sites taht are biallelic, for now assume ISM
more<-basic@n.polyallelic.sites
varsite<-Nsite+more

filehaps <- F_ST.stats(file,mode="haplotype",only.haplotype.counts=TRUE)
haplotypecounts <- filehaps@region.stats@haplotype.counts
# this is helpful https://github.com/cran/PopGenome/blob/master/vignettes/Integration_of_new_Methods.Rnw 

########

#install.packages("entropart")
library(entropart)
#need frequencies of haplotypes reported from PopGenome so you can use entropart to get 2D
hapfreq<-unlist(haplotypecounts)
actual<-sum(hapfreq)
hapfreq<-hapfreq/actual
numhaps<-length(hapfreq)

Simp<-expq(Simpson(Ps=hapfreq),q=2)
Gini<-1-(1/Simp)

# Watterson estimator of theta
harm=0
for (n in 1:(actual-1)){
  harm=harm+1/n
}

QW <- varsite/harm

```

```{r hapdivexample, echo=FALSE,warning=FALSE,results='hide',message=FALSE,fig.show='asis'}

x=1
cdf=0
indprob=0
array<-NULL
while (cdf<0.95) {
  cdfprev<-cdf
  cdf<-pgamma(x,numhaps,Hapdiv) #might be that numhaps is actually the shape parameter!!!! or: something else...non-gamma.
  indprob<-cdf-cdfprev
  
  
  happrob<-numhaps+(x-1)
  array<-c(array,happrob)
  array<-c(array,cdf)
  array<-c(array,indprob)
  x=x+1
}
min=numhaps
max=x
probs<-t(matrix(array,nrow=3))
probs
#plot(probs[,1],probs[,3],col='red')

```


Evaluating a sample of sequences from the barnacle *Notochthamalus scabrosus*, where `r actual` individuals were haphazardly sampled from the data of Laughlin et al. [-@Laugh12], we see that these data would traditionally report haplotype diversity *H* of `r Hapdiv`, from `r numhaps` observed haplotypes (most dominant haplotype at frequency `r max(hapfreq)`), and in this instance the Gini-Simpson index is equal to `r Gini`. Now we have another statistic that can be calculated from previous data on the population, that focuses on the number of dominant haplotypes. Here, the inverse Simpson index $^2D$ is `r Simp`, the effective number of haplotypes in the system. Using the approach detailed above, but given these data and assuming that the number of input sequences is unknown, we could say there are between `r min` and `r max` individuals that went into the sample.

This model is intuitively useful - the number of individuals cannot be less than the number of haplotypes, and is likely more; the number of individuals could be much more than the number of haplotypes if marker diversity is low; and the distribution scales up as the minimum increases - however is mathematically not well-informed. As noted above, there are specific probability distributions associated with a sample of sequences, the number of segregating sites *S*, and a prior assumption of $\theta$ [@Wakeley08].

$$ P(S=k) = \sum\limits_{i=2}^{n}(-1)^i{n-1 \choose i-1}\frac{i-1}{\theta+i-1}(\frac{\theta}{\theta+i-1})^k $$

Figure Xa illustrates this distribution for $\theta$=2. This represents a low-diversity population, and unless few segregating sites are observed there may be a broad range of sample sizes consistent with such an observation. Figure Xb illustrates the same probability distribution, but assuming $\theta$=10. When the prior knowledge or assumption of diversity is higher, there tends to be a sharper distribution on *n* for a given *k*.


```{r WakeleyCh4, echo=FALSE,warning=FALSE,results='hide',message=FALSE,fig.show='asis'}
library(plot3D)
Q = 2 # this is PRIOR information from what you know of input SPECIES, not that particular data set (N. scab Q about 10)
maxn = 20 #above 70 this behaves funny???? OR IT MAY BE SOME MULTIPLE OF Q*maxn that is problem?

obsvdk = varsite
maxk = 15 #must be greater than obsvdk



a <- c(1:maxn)
b <- c(1:maxk)
PSKv <- numeric(maxn*maxk)
PSKv <- matrix(PSKv,ncol=maxn)
colnames(PSKv)<-a
rownames(PSKv)<-b

for (n in 2:maxn) {
#  print ("n")
#  print (n)
  for (k in 0:maxk) {
#    print (k)
    PSK=0
    for (i in 2:n) {
#      print (i)
      PSK<-PSK + ((-1)^i)*(choose((n-1),(i-1)))*((i-1)/(Q+i-1))*(Q/(Q+i-1))^k
      
    }
#    print (PSK)
    PSKv[k,n]<-PSKv[k,n]+PSK
    obsvd<-PSKv[obsvdk,]

  }
  
}

Q=10
a2 <- c(1:maxn)
b2 <- c(1:maxk)
PSKv2 <- numeric(maxn*maxk)
PSKv2 <- matrix(PSKv2,ncol=maxn)
colnames(PSKv)<-a2
rownames(PSKv)<-b2

for (n in 2:maxn) {

  for (k in 0:maxk) {

    PSK2=0
    for (i in 2:n) {

      PSK2<-PSK2 + ((-1)^i)*(choose((n-1),(i-1)))*((i-1)/(Q+i-1))*(Q/(Q+i-1))^k
      
    }

    PSKv2[k,n]<-PSKv2[k,n]+PSK2


  }
  
}


par(mfrow=c(1,2))

hist3D(x=seq(0,1,length.out=nrow(PSKv)),y=seq(0,1,length.out=ncol(PSKv)),PSKv,col=NULL,border="black",theta=30,phi=30,xlab="k",ylab="n",zlab="P{S=k}",alpha=0.35)

hist3D(x=seq(0,1,length.out=nrow(PSKv2)),y=seq(0,1,length.out=ncol(PSKv2)),PSKv2,col=NULL,border="black",theta=30,phi=30,xlab="k",ylab="n",zlab="P{S=k}",alpha=0.35)

par(mfrow=c(1,1))

```

We propose two/three methods that could help to estimate the relative number of individuals in a barcoding sample: 1) gamma distribution, 2) theta and segregating sites, 3) sampling theory?. To evaluate the usefulness of each method, we used a simulation approach. We simulated populations with Hudson's MS program [@Hudson 2002] using the *gap* package in R [@RCoreTeam2013]. We simulated four populations, using two different mutation rates (theta= 2 and theta=10), and also considering populations with or without growth. The growth rate was defined to achieve the Tajima's D value that better approximates the values observed in nature for marine invertebrates. Detailed information on simulations and R code is presented in the supplementary material.

```{r simulating source populations, results='asis',echo=FALSE}
library(gap)
library(PopGenome)
library(knitr)

# -----gap----
# we can call ms from R without having to use the console and read the object directly with the gap function

# with -G x we set a population growth, a negative value of x means that the population was larger in the past that at present. 

# Simulating our source populations
theta2Growth <- system("ms 1000 1 -t 2 -G 10", intern=TRUE)
theta2NoGrowth <- system("ms 1000 1 -t 2", intern=TRUE)
theta10Growth <- system("ms 1000 1 -t 10 -G 10", intern=TRUE)
theta10NoGrowth <- system("ms 1000 1 -t 10", intern=TRUE)

# we need to write the outputs so we can read them using PopGenome
#write(theta2Growth,"theta2Growth.out")
#write(theta2NoGrowth,"theta2NoGrowth.out")
#write(theta10Growth,"theta10Growth.out")
#write(theta10NoGrowth,"theta10NoGrowth.out")

# reading with PopGenome to calculate H and Tajima's D for each of the four populations
readMS("theta2Growth.out")->popgen.t2g #gives a "genome" object
readMS("theta2NoGrowth.out")->popgen.t2ng
readMS("theta10Growth.out")->popgen.t10g
readMS("theta10NoGrowth.out")->popgen.t10ng

# run F_ST stats and check haplotype diversity
F_ST.stats(popgen.t2g)->popgen.t2g
F_ST.stats(popgen.t2ng)->popgen.t2ng
F_ST.stats(popgen.t10g)->popgen.t10g
F_ST.stats(popgen.t10ng)->popgen.t10ng

# get haplotype diversity for each source population
unlist(popgen.t2g@region.stats@haplotype.diversity)->hapDiv.t2g
unlist(popgen.t2ng@region.stats@haplotype.diversity)->hapDiv.t2ng
unlist(popgen.t10g@region.stats@haplotype.diversity)->hapDiv.t10g
unlist(popgen.t10ng@region.stats@haplotype.diversity)->hapDiv.t10ng

hapDiv<-c(hapDiv.t2g,hapDiv.t2ng,hapDiv.t10g,hapDiv.t10ng)
hapDiv

# run neutrality.stats to check that Tajima's D is negative and looks real
neutrality.stats(popgen.t2g)->popgen.t2g
neutrality.stats(popgen.t2ng)->popgen.t2ng
neutrality.stats(popgen.t10g)->popgen.t10g
neutrality.stats(popgen.t10ng)->popgen.t10ng

Taj.t2g<-popgen.t10g@Tajima.D
Taj.t2ng<-popgen.t10ng@Tajima.D
Taj.t10g<-popgen.t10g@Tajima.D
Taj.t10ng<-popgen.t10ng@Tajima.D

tajima<-c(Taj.t2g,Taj.t2ng,Taj.t10g,Taj.t10ng)
tajima

# summary table
thetas<-c(2,2,10,10)
pops<-c("Population 1", "Population 2", "Population 3", "Population 4")
growth<-c("yes","no","yes","no")
newhap<-round(hapDiv,2)
newtaj<-round(tajima,2)
sumTable<-cbind(pops,thetas,growth,newhap,newtaj)
colnames(sumTable)<-c("Population","Theta","Growth","Haplotype diversity","Tajima's D")
kable(sumTable)
```

From these four populations, we took "field samples" of different sizes (n), sampling without replacement. We replicate the sampling experiment 100 times, to be able to assess variation of sampling. For each replicate, we calculate the number of haplotypes and the number of segregating sites, which represent our observed values in our simulated samples. The sampling size, that it is known to us from this design, is what we are going to try to predict using the back calculations described above. All the analysis of the simulated populations was done in R [@RCoreTeam2013].

```{r taking field samples}
library(gap)
library(PopGenome)

# reading the ms output generated in the previous chunk with gap
# this way works for reading the object from R, but it is not useful if we want to maintain the origina same populations, because I was not able to make the "seeds" work
#read.ms.output(theta2Growth,FALSE)->pop1
#read.ms.output(theta2NoGrowth,FALSE)->pop2
#read.ms.output(theta10Growth,FALSE)->pop3
#read.ms.output(theta10NoGrowth,FALSE)->pop4

# this way, we read the same files that were simulated only once, and the same that we read with PopGenome
read.ms.output("theta2Growth.out",is.file=T)->pop1
read.ms.output("theta2NoGrowth.out",is.file=T)->pop2
read.ms.output("theta10Growth.out",is.file=T)->pop3
read.ms.output("theta10NoGrowth.out",is.file=T)->pop4

# --------sampling from each population---------
# vector with all the populations for loops or easy coding
pop<-list(pop1,pop2,pop3,pop4)

# vector with the sizes we want to sample
#popsizes<-c(2,4,8,16,32,64,128)
#popsizes<-seq(1:128)

# we extract the gametes matrix for each population, from there we take the samples, we put the matrix obtained in each "sample" in a list called "samples", and the result for each population in a list called "populations"

set.seed(123) #set seed for random sampling

# loop in each population to take the field samples, 100 replicates for each sampling size
populations<-list() #empty list to put results
for(i in 1:length(pop)){
  pop[[i]]->ourpop #pick a population
  namepop<-paste("popOrigen",i,sep="")
  t(ourpop$gametes[[1]])->mat #transpose matrix to have individuals as rows
  samples<-list() #empty list to put the sampling of each sample size
    for(j in popsizes){
        name<-paste("sampleSize",j,sep="")
        replicates<-list() #empty list to put each replicate (n=100)
        for (k in 1:100){
        namerep<-paste("replicate",k,sep="")  
        mat[sample(nrow(mat),size=j,replace=FALSE),]->rep
        rep->replicates[[namerep]]
        }
        replicates->samples[[name]]
      }
  samples->populations[[namepop]]
} # now in each of the 4 components of the list "populations", we have 7 list with each component corresponding to each sample size, and within each of this list we have 100 components corresponding to each replicate of the sampled matrix.
```

```{r calculating haplotypes}
# Now for each sample we need to calculate the segregating sites and number of haplotypes
# the haplotype number can be extracted by doing unique() of the matrix
hapResults<-list() #create list to save the results
for(i in 1:length(pop)){
  populations[[i]]->popnow #choose population to work
  namepop<-paste("popOrigen",i,sep="")
  samples<-list() #create list to put each sampling size
  for (j in 1:length(popsizes)){
    namesample<-paste("sampleSize",j,sep="")  
    popnow[[j]]->samplenow #choose sampling size to work with
    print(paste("working in:",namepop,namesample))
    for (k in 1:100){
      sapply(samplenow,function(x) nrow(unique(x)))->result
      unlist(result)->replicates
      }
    replicates->samples[[namesample]]
    }
  samples->hapResults[[namepop]]
} # now hapResults is a list of vectors with the haplotypes number for each population, in each sample size
hapResults[[1]]->pop1haps
hapResults[[2]]->pop2haps
hapResults[[3]]->pop3haps
hapResults[[4]]->pop4haps
```
We are going to do a similar procedure to estimate the number of segregating sites. We first created a function to estimate the number of segregating sites from the gametes matrix provided by the MS simulation. And the we apply that looping over our samples.

```{r calculating segregating sites}
# load function to estimate segregating sites, it works taking a matrix of gametes and returning a numeric value, the number of segregating sites.
estimate.segSites <- function(myMatrix){
  answers<-rep(NA,(nrow(myMatrix)-1))
  sites<-rep(NA,ncol(myMatrix))
    for (j in 1:ncol(myMatrix)){
      for (i in 1:(nrow(myMatrix)-1)){
        (myMatrix[i+1,j]==myMatrix[1,j])->answers[i]
      }
      if (all(answers)==TRUE) {0->sites[j]}
      else {1->sites[j]}
      }
    sum(sites,na.rm=T)->segSites
    return(segSites)
  }

# to get segsites for each population we need to use our custome function in a loop
segResults<-list()
for(i in 1:length(pop)){
  populations[[i]]->popnow
  namepop<-paste("popOrigen",i,sep="")
  samples<-list()
  for (j in 1:length(popsizes)){
    namesample<-paste("sampleSize",j,sep="")
    popnow[[j]]->samplenow
    sapply(samplenow,function(x) estimate.segSites(x))->result
    result->samples[[namesample]]
    }
  samples->segResults[[namepop]]
}  # now segResults is a list of 4 lists (one for each population), within each of those list compose of a list of 7 vectors, one for each sample size, and with a length of 100, where each component is the segregating sites for the 100 replicates. 
segResults
segResults[[1]]->pop1segsites
segResults[[2]]->pop2segsites
segResults[[3]]->pop3segsites
segResults[[4]]->pop4segsites
```

Since these calculations take quite a while (at least the haplotypes), we are going to save our results in a dataframe, the number of haplotypes and the number of segreating sites for each population. The following chunk may not be very efficient but it works fine.


5-3-15 we are now pretty sure that the S=k method is too insensitive (as often the case iwth coalescent, and we see that above samples of 10 the values are relatively insensitive, see Vince Buffalo's "dancing genealogies" for illustration of why even if peaks look sharper for theta 10 the possibilities are massive for the empirical stuff that fits....). So this is concerning for even being able to tell orders of magnitude apart, e.g. 10 of species A and 100 of species B probably won't happen. 

So new structure of paper we think: show that haplotypes are the defined minimum number of individuals, this is the dumb but easy and factual way. Then the gamma distribution, which is a shot in the dark, makes sense but parameterization is hard to think about. Then the 'smart' coalescent way, but we figure out it is not very sensitive.

And then part B of Results would be that we show empirical values of #haplotypes for samples of n given theta, identify whether this helps us parameterize the slightly-smarter approach. We may also ponder the mathematical relationship between theta and hapdiv, but of course for a given theta (truth) there is a range of theta (estimated from empirical) and a range of hapdiv (estimated from empirical)



# Results 

General information of the simulated populations is presented in Table 1. 

```{r nextfig, echo=FALSE,warning=FALSE,results='hide',message=FALSE,fig.show='asis'}
#  print(obsvd) #now would be easy to print 95% HPD of obsvd, so for given theta and k this is the distribution of n
  plot(obsvd,xlab="n",ylab="P(n|k)",ylim=c(0,0.15),main=paste("for K =",obsvdk, "and theta =",Q,"in black; for hapdiv in red"))
points(probs[,1],probs[,3],col='red')
abline(v=actual)


```

Remember that the input data for this single taxon included `r actual` individuals, the vertical line in plot above. What is likelihood function? Product of the two distributions? That is too stark in areas where they don't really overlap probabilities. Shouldn't be ZERO there?

Need to then adapt this to a sample from 4-5 species for which there is known information? Or do simulated data (X species, vector of theta in simulation and hapdiv when all said and done, they are after all, related) and adjust the evenness in series of plots to see if evenness/richness gets recovered appropriately, given confidence intervals after all...

n.b. Marc Feldman was concerned about the two statistics double-dipping on the same theory... is this ABC, is this borrowing strength, is this inappropriate?? **AH. but certainly *S* and *H* are not independent observations thus cannot simply use product of the two to generate a Likelihood** perhaps identify which is better, useful, or are neither sufficient?


# Discussion

Returning to the coin flip, it is worth evaluating wherein lies the strength of inferential signal. 50% gives NO information, could be 2 or infinite flips. So it is deviation that is signal in the coin flip example. Similarly, for a system of diversity such as this we need the *potential* for diverse outcomes to evaluate: low theta means nearly all sample sizes are possible, for example. In this sense, developing this with a mind for species that are broadly distributed and highly abundant is likely a more effective strategy than endemic small populations.

Talk a bit about how barcode frequency information maybe isn't as far off when dealing with closely related taxa? We aren't throuwing out frequency, goal here is to look solely at complementary information.

In the end, Discussion :: though there are concerns about read frequency...at a minimum haplotpye number bounds the minimum. In this way we feel better about order of magnitude results, and less need for prior information from a population. This has been less evaluated in microbial/viral samples as "population" is perhaps less defined in those communi9ties than in eukaryote or metazoan communiteis, e.g. through gene flow. But as a complementary recognition that the number of haplotypes tells us some information for sure, and prior information about that population may also provide additional information, we may start to improve on our ability to recover actual ecology from actual molecules.

5-3-15 we are now pretty sure that the S=k method is too insensitive (as often the case iwth coalescent, and we see that above samples of 10 the values are relatively insensitive, see Vince Buffalo's "dancing genealogies" for illustration of why even if peaks look sharper for theta 10 the possibilities are massive for the empirical stuff that fits....). So this is concerning for even being able to tell orders of magnitude apart, e.g. 10 of species A and 100 of species B probably won't happen. 

# Acknowledgments

Idea brought about by extended problem-solving session with J. Drake, helped greatly by C. Ewers-Saucedo and K. Bockrath. Work supported by funding from NSF-OCE-Chile, OVPR, and UGA Department of Genetics.

# Literature Cited


