---
title: "Doing Simulations"
author: "Paula Pappalardo"
date: "Wednesday, April 22, 2015"
output: pdf_document
---

# Running simulations

After the first trials, I realized the best method is to call "ms" from R, read that from *gap* and save the file to read it also with *PopGenome*. With *gap* we are able to estimate number of haplotypes. With *PopGenome* we are able to estimate haplotype diversity and also Tajima's D for the cases that we want to simulate "real" populations.

We are going to simulate four populations of 1000 individuals, with the following parameters:

1) theta=2, growth rate to have "good" Tajima's D
2) theta=10, growth rate to have "good" Tajima's D
3) theta=2, no growth
4) theta=10, no growth.

For these four populations we are going to calculate the haplotype diversity and the Tajima's D (to check).

From each of the four populations we are going to sample without replacement "field samples" of 2,4,8,16,32,64 and 128 individuals. That way we know the real n. Using the number of haplotypes and the number of segregating sites of the "field samples" we are going to do our best to back calculate the number of individuals in that sample, using as prior information the haplotype diversity of the original population (n=1000).

The only thing that we need to keep from each "field sample" is the number of haplotypes and number of segregating sites. So to save memory I may try to clean the files we are not using if things get problematic...

NOTE: one thing to have in mind is that we are not controlling for sequence length since there are no sequences in the ms output and we are just seing the segregating sites. That's why the sequence length that is usually visibly with the summary stats in PopGenome is zero when we are reading the ms output file. It shouldn't matter for the things we discussed.

Let's do it! :)

```{r simulating source populations}
library(gap)
library(PopGenome)

# -----gap----
# we can call ms from R without having to use the console and read the object directly with the gap function

# with -G x we set a population growth, a negative value of x means that the population was larger in the past that at present. 

# Simulating our source populations
theta2Growth <- system("ms 1000 1 -t 2 -G 10", intern=TRUE)
theta2NoGrowth <- system("ms 1000 1 -t 2", intern=TRUE)
theta10Growth <- system("ms 1000 1 -t 10 -G 10", intern=TRUE)
theta10NoGrowth <- system("ms 1000 1 -t 10", intern=TRUE)

# we need to write the outputs so we can read them using PopGenome
write(theta2Growth,"theta2Growth.out")
write(theta2NoGrowth,"theta2NoGrowth.out")
write(theta10Growth,"theta10Growth.out")
write(theta10NoGrowth,"theta10NoGrowth.out")

# reading with PopGenome to calculate H and Tajima's D for each of the four populations
readMS("theta2Growth.out")->popgen.t2g #gives a "genome" object
readMS("theta2NoGrowth.out")->popgen.t2ng
readMS("theta10Growth.out")->popgen.t10g
readMS("theta10NoGrowth.out")->popgen.t10ng

# run F_ST stats and check haplotype diversity
F_ST.stats(popgen.t2g)->popgen.t2g
F_ST.stats(popgen.t2ng)->popgen.t2ng
F_ST.stats(popgen.t10g)->popgen.t10g
F_ST.stats(popgen.t10ng)->popgen.t10ng

# get haplotype diversity for each source population
unlist(popgen.t2g@region.stats@haplotype.diversity)->hapDiv.t2g
unlist(popgen.t2ng@region.stats@haplotype.diversity)->hapDiv.t2ng
unlist(popgen.t10g@region.stats@haplotype.diversity)->hapDiv.t10g
unlist(popgen.t10ng@region.stats@haplotype.diversity)->hapDiv.t10ng

hapDiv<-c(hapDiv.t2g,hapDiv.t2ng,hapDiv.t10g,hapDiv.t10ng)
hapDiv

# run neutrality.stats to check that Tajima's D is negative and looks real
neutrality.stats(popgen.t2g)->popgen.t2g
neutrality.stats(popgen.t2ng)->popgen.t2ng
neutrality.stats(popgen.t10g)->popgen.t10g
neutrality.stats(popgen.t10ng)->popgen.t10ng

Taj.t2g<-popgen.t10g@Tajima.D
Taj.t2ng<-popgen.t10ng@Tajima.D
Taj.t10g<-popgen.t10g@Tajima.D
Taj.t10ng<-popgen.t10ng@Tajima.D

tajima<-c(Taj.t2g,Taj.t2ng,Taj.t10g,Taj.t10ng)
tajima
```

Now that we have our 4 populations and we have H and Tajima's D for those, we need to take our "field samples". For that, we are going to use the same pop generated above and read those with gap. Then, using the matrix of the variables sites for the 1000 individuals we are going to take random samples. We are going to use sample sizes from 2 to 128 in a log2 scale.

```{r taking field samples}
library(gap)
library(PopGenome)

# reading the ms output generated in the previous chunk with gap
read.ms.output(theta2Growth,FALSE)->pop1
read.ms.output(theta2NoGrowth,FALSE)->pop2
read.ms.output(theta10Growth,FALSE)->pop3
read.ms.output(theta10NoGrowth,FALSE)->pop4

# --------sampling from each population---------
# vector with all the populations for loops or easy coding
pop<-list(pop1,pop2,pop3,pop4)

# vector with the sizes we want to sample
popsizes<-c(2,4,8,16,32,64,128)

# we extract the gametes matrix for each population, from there we take the samples, we put the matrix obtained in each "sample" in a list called "samples", and the result for each population in a list called "populations"

populations<-list()

for(i in 1:length(pop)){
  pop[[i]]->ourpop
  namepop<-paste("popOrigen",i,sep="")
  t(ourpop$gametes[[1]])->mat
  samples<-list()  
    for(j in popsizes){
        name<-paste("sampleSize",j,sep="")
        mat[sample(nrow(mat),size=j,replace=FALSE),]->sampMat
        sampMat->samples[[name]]
      }
  samples->populations[[namepop]]
}  

# now in each of the 4 component of the list "populations", we have the 7 list with the samples corresponding to each of the sampling sizes 
```

Now we are going to calculate number of  haplotypes and number of segregating sites for each of the samples in each of the 4 populations.

```{r calculating haplotypes and segsites for samples}
# Now for each sample we need to calculate the segregating sites and number of haplotypes

# the haplotype number can be extracted by doing unique() of the matrix
hapResults<-list()
for(i in 1:length(pop)){
  populations[[i]]->popnow
  namepop<-paste("popOrigen",i,sep="")
  sapply(popnow,function(x) nrow(unique(x)))->haplotypes
  haplotypes->hapResults[[namepop]]
} # now hapResults is a list of vectors with the haplotypes number for each population, in each sample size

# to get segsites for each population, we can count the number of segregating sites by counting the colummns of our sampled matrix
segsitesResults<-list()
for(i in 1:length(pop)){
  populations[[i]]->popnow
  namepop<-paste("popOrigen",i,sep="")
  sapply(popnow,function(x) ncol((x)))->segsites
  segsites->segsitesResults[[namepop]]
}

```

